Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-09-29 paper.tex
more edits

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios are compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator is used as a level playing field to compare
the scenarios.  When under supply-constraint pressure, the four cases exhibit
noticably different behavior.  Otherwise, they generally behaved very
similarly.  Fleet-based modeling is more efficient in supply-constrained
environments at the expense of losing insight on issues such as realistically
sub-optimal fuel distribution and challenges in reactor refueling cycle
staggering.  Finer-grained time steps also enable more efficient material use
in supply-constrained environments.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility modeling decisions. Several
effects are described below that result from fundamental differences between
these two modeling paradigms.

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short a half batch of fuel it will produce $\frac{5}{6}\cdot C$ of
power for that time step where $C$ is the power capacity of a single reactor
of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is short a
half batch of fuel, then it will produce no power that time step.  Time step
duration can also have this effect.  For both fleet and individual reactor
modeling, power goes offline in full time step increments.  For example, if a
3-batch fleet based reactor is short $\frac{1}{2}$ batch of fuel with a 3
month time step, then it will produce $\frac{5}{6} \cdot 3 \text{mo} \cdot C $
of energy.  In contrast, a monthly time step might allow the same reactor to
produce $[\frac{5}{6} \cdot 1 \text{mo} + 2 \text{mo}] \cdot C $ of energy if
the fuel shortage actually ended after only 1 month.  Quantifying this effect
directly is difficult.  One way of observing it is to compare fleet reactor
scenarios with that have different time step durations (i.e. cases 2 and 4).
Because the fleet reactors have perfect fuel sharing (described below) and no
outage noise, differences observed will likely be due to quantized shutdown.

Individually modeled reactors each have their own refueling cycles.  Depending
on how the refueling cycles and outages are staggered, power production can
vary significantly.  For example, if all reactors refuel at the same time,
then all reactors would be offline at the same time.  This is referred to as
the \emph{cycle staggering effect} and does not occur in fleet-based reactor
models.  An example of poor cycle staggering is shown in Figure
\ref{fig:sync-cycle}.  Poor staggering causes spikes of material supply and
demand that can lead to suboptimal resource utilization.  A fuel fabrication
facility might receive requests for new batches for every reactor all at once.
The reactors draw out inventory from the fuel fabrication facility together in
a large quantity of orders.  If the fabrication facility does not have enough
on hand, some number of reactors will need to wait until enough fuel can be
pumped through fabrication.  Avoiding such constraints would require the
fabrication facility to maintain suboptimally large on-hand inventories as a
contingency.  This effect can be naturally observed as the fluctuations in the
difference between deployed power capacity and generated power.

>>>>

Due to the diversity of fuel cycle simulators' modeling assumptions direct
comparison and benchmarking can be difficult.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study that
is perhaps one of the most complete published comparisons performed.  Despite
this, however, various results from the simulators were often significantly
different due to inconsistency in modeling decisions involving reprocessing
strategies, refueling behavior, reactor end-of-life handling, etc.  This work
attempts to identify and quantify the effects of a few different modeling
choices that may sometimes be taken for granted in the fuel cycle modeling
domain.  Four scenarios are compared using combinations of either fleet-based
or individual reactors with either monthly or quarterly (3-month) time steps.
The Cyclus fuel cycle simulator is used as a level playing field to compare
the scenarios.  When under supply-constraint pressure, the four cases exhibit
noticably different behavior.  Otherwise, they generally behaved very
similarly.  Fleet-based modeling is more efficient in supply-constrained
environments at the expense of losing insight on issues such as realistically
suboptimal fuel distribution and challenges in reactor refueling cycle
staggering.  Finer-grained time steps also enable more efficient material use
in supply-constrained environments.  Large simulations with fleet-based
reactors run much more quickly than their individual reactor counterparts.

@@

There are various low level phenomena that have potential to affect the larger
outcome of simulations with respect to facility and time step modeling
decisions. Several effects are described below that arise from these
different modeling choices.

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot C$
of power for that time step where $C$ is the power capacity of a single
reactor of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot C$) that
time step.  Time step duration can also have this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
For example, if a 3-batch fleet based reactor is short $\frac{1}{2}$ batch of
fuel with a 3 month time step, then it will produce $\frac{2.5}{3} \cdot 3
\text{mo} \cdot C $ of energy.  In contrast, a monthly time step might allow
the same reactor to produce $[\frac{5}{6} \cdot 1 \text{mo} + 1 \cdot 2
\text{mo}] \cdot C $ of energy if the fuel shortage actually ended after only
1 month.  Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases 2 and 4).  Because the fleet reactors have perfect fuel
sharing (described below) and no outage noise, differences observed will
likely be due to quantized shutdown.

Individually modeled reactors each have their own refueling cycles.  Depending
on how the refueling cycles and outages are staggered, power production can
vary significantly.  For example, if all reactors refuel at the same time,
then all reactors are offline at the same time.  This is referred to as the
\emph{cycle staggering effect} and does not occur in fleet-based reactor
models.  An example of poor cycle staggering is shown in Figure
\ref{fig:sync-cycle}.  This effect can be naturally observed as the
fluctuations in the difference between deployed power capacity and generated
power.

Poor reactor cycle staggering can also cause spikes of material supply and
demand that can lead to suboptimal resource utilization.  Consider a fuel
fabrication facility that receives requests for new batches for every reactor
all at once.  The reactors draw out inventory from the fuel fabrication
facility together in a large quantity of orders.  If the fabrication facility
does not have enough material on hand, some number of reactors will need to
wait until enough fuel can be pumped through fabrication.  Avoiding such
constraints would require the fabrication facility to maintain suboptimally
large on-hand inventories as a contingency.  Even with infinite material
availability, poor staggering can result in supply constraints caused by
finite facility throughput. Even if the fuel fabrication facility above has
infinite material supply that it can keep on hand, it might not be able to
fabricate fuel for all reactor requests all at once even though it has
sufficient capacity if they request fuel uniformly over time.
